Water desalination using polyelectrolyte hydrogel: Gibbs ensemble modeling

### 2022-09-02 gibbs.tex
Update gibbs.tex

\paragraph{Water desalination technologies.}

@@

\paragraph{Hydrogels for desalination.}
Hydrogels are three-dimensional networks of polymer chains that are crosslinked by either physical or chemical bonds. 
They are able to entrap large volumes of water attracted by the high concentration of hydrophilic groups. 
When a dehydrated or deswollen hydrogel uptakes water, its polymer chains extend, creating a swelling pressure. 
For example, as reported in \cite{Wack_2009} weakly crosslinked poly(acrylic acid) (PAA) copolymers with polymer volume fractions between 0.03 and 0.30 exhibit a swelling pressure ranging from 0.20--4.23 MPa.
Polyelectrolyte hydrogels, which are carrying ionic groups on the comonomer units (like PAA), are able to reject salt ions from the solution, they absorb a solution of lower salinity than that they are equilibrated with.

@@

\paragraph{Physics behind the desalination process.}
%The maximum entropy principle requires the density of charges to be more-less the same throughout the system medium. 
Since the polyelectrolyte gel has its own charges and its own neutralizing counterions, the density of mobile ions (which are able to freely enter and leave the gel) in gel appears to be smaller than their density outside the gel.
Therefore the internal solution of the gel has a lower density of mobile ions than that in the solution outside. 
In that sense, the gel acts as an osmotic membrane separating solutions \cite{Rud2020}. 
The driving force of the separation is the Donnan potential originated from the charges of the hydrogel network. 
The density of mobile ions difference between ions concentrations in the internal and external solutions
is defined by Donnan law \cite{Rud2018}
\begin{equation*}
\frac{\ccl\gel}{\cs}=\frac{\cs}{\cna\gel}=\sqrt{1+\left(\frac{\alpha}{2\cs\Vgel}\right)^{2}}\pm\frac{\alpha}{2\cs\Vgel}
\end{equation*}
where $\ccl\gel$ and $\cna\gel$ are the internal concentration of monovalent anions and cations, $\cs$ is the salinity of external solution,  $\Vgel$ is the gel molar volume (inverse density of gel segments) and $\alpha$ is ionization degree (in the case of our study $\alpha=1$). 
The $\pm$ sign in the formula accounts for the change between the cases of gel made of polyanion or polycation.

@@

%Let $\nna$ and $\ncl$ be the density of the corresponding ion in the compression volume, $\Vbox$ averaged over this volume
\begin{equation}
N_i = N_i\gel + \cs\cdot(\Vbox - \Vgel)
\end{equation}
where $i \in \{\na, \cl\}$,  $N_i$ is the amount of the corresponding ions in gel, $\Vgel$~--- volume of the gel, $\Vbox$~--- the volume in which the gel is compressed. 
As $\Vbox$ we choose the volume of the gel in the free swelling equilibrium state.

%Then in closed system $\nna$ and $\ncl$ remain constant, whereas the salinity of supernate, $\cs$, may change with the change of the gel volume.

\paragraph{Open system.}
We simulate the open system using the grand reaction method \cite{Landsgesel2020a,Rud2020}.
This method is a hybrid of molecular dynamics (MD) and Monte Carlo (MC). 
The whole simulation represents a chain of subsimulations of MD and MC followed one by each other. 
For MD simulation we used a standard Langevin dynamics \cite{Grest1986} which models the mechanical movement of all the system particles. 
Whereas MC simulates the thermodynamic equilibrium with reservoir, which exchanges ions with the simulation box. 
The insertion (and deletion) of ion pairs is considered as a reaction of creation (or annihilation) of an ion pair 
\[
\varnothing\stackrel{K}{\leftrightarrows}\na+\cl
\]
with a reaction constant defined by the chemical potential of ions,
$K=\exp\left(\muna+\mucl\right).$

\paragraph{Closed system.}
The similar scheme we use to simulate the gel in closed system. 
The procedure represents a chain of subsequent MD and MC simulation, but now the molecular dynamics is running in two simulation boxes simultaneously.
The first box of the volume $\Vgel$ contains the gel with ions, and the second box, $\Vout$--- contains only ions. 
The total volume of both boxes is kept constant, $\Vbox=V_{\mathrm{gel}}+V_{\mathrm{out}}$, so the compression of the gel implies the decreasing $\Vgel$ and increasing $\Vout$.
These two boxes are represented in \reffig{fig:diamond}.

\paragraph{Algorithm.}
The thermodynamic equilibrium between the two volumes implies the exchange of ion pairs between them, which is done by means of MC procedure in a way similar to described in \cite{Panagiotopoulos1988b}.
The simulation algorithm is following:
\begin{enumerate}
\item Simulate molecular dynamics in both boxes; collect the observables, for example, values of pressure $P$, to an array; 
\item Perform Monte-Carlo procedure simulating ion pair exchange; collect the arrays of observables samples: number of ions in both boxes, 
%$\nna\gel$,
$\Ncl\gel$ 
%$\nna\out$ 
and $\Ncl\out$ ; 
\item Repeat the procedure until the desired length of sample arrays is reached.
\end{enumerate}

@@

\paragraph{Compression in open system.}
\begin{figure*}[t]
\subfloat[The gel partial pressure vs gel molar volume \label{fig: PV}]
{\includegraphics[width=0.48\textwidth]{figures/fig_PV_pub.pdf}}
\hspace{0.02\textwidth}
\subfloat[Supernate salinity vs gel molar volume\label{fig: CV}]
{\includegraphics[width=0.48\textwidth]{figures/fig_CV_pub.pdf}}
\caption{
The compression of the gel in open system (solid lines) and in closed system (dotted lines). 
Each solid curve corresponds to different salinity of the reservoir $\cs$ (see legend). Shadowed area limit the states with applied pressure below zero and above 5 bar.\label{fig: PV and CV}}
\end{figure*}
First, we run a set of simulations modeling the gel compression in \emph{open system}, \ie in equilibrium with a big bath of certain salinity, $\cs$. 
The simulations are run for a set of different simulation box volumes. 
Each simulation returns the averaged pressure, $P$, and the number of $\cl$ ions, which are present in the simulation box, $\ncl$.

@@

\paragraph{Compression in closed system.}
%Each set of simulations in \emph{open system} defines the free swelling equilibrium states of the gel for each salinity, \ie the state where the gel partial pressure equals zero. 
The free swelling equilibrium states are characterized by the corresponding molar volume of the gel, $\Vgel^0$ and the amount of ions in gel \{$\Nna^0$, $\Ncl^0$\} 
(Index ``0'' stands for zero bar applied pressure).
We use these states as the starting ones for the gel compression in \emph{closed system}. 
Thus, we simulate the compression of the gel in a closed volume, $\Vbox$, which is equal to $\Vgel^0$, and contains $\Ncl^0$ of $\cl$ ions and the $\na$ ions in the amount neutralizing the system, $\Nna^0 = $$\Ncl^0 + \Ngel$.

@@

\paragraph{Desalination scheme.}
As we have seen, the compression of the gel in a closed system affects the salinity, whereas the compression in an open system affects the amount of ions in a volume where the gel is compressed in. 
Here we show how to employ these phenomena for water desalination. 
The highlighted colored lines on plots of Figures \ref{fig: PV and CV} and \ref{fig: NV and CN} form a sequence of the following one by each other swellings and compressions of the hydrogel, correspondingly in \emph{open} and \emph{closed} systems. 
This sequence forms the water desalination process.
Starting from swelling the gel in the \emph{open} system at high salinity ($\cs=0.091$ mol/l, solid black line), the gel gets compressed in \emph{closed} system until the pressure reaches 5 bar (dashed black line). 
Then the same gel swells in a reservoir of a bit lower salinity (\ie $\cs=0.064$ mol/l, light blue line). 
After swelling, the gel is compressed again by the pressure of 5 bar in \emph{closed} system (dashed light blue line).  
Then the gel is put swelling in a reservoir of even smaller salinity ($\cs=0.045$ mol/l, solid yellow line). 
And so on. 
The chain of alternating swelling and compression steps ends up when salinity gets equal to $\cs=4\cdot10^{-3}$ mol/l after compression in closed system (dashed magenta line).

>>>>

\paragraph{Water desalination technologies}

@@

\paragraph{Hydrogels for desalination}
Hydrogels are three-dimensional networks of polymer chains that are crosslinked by either physical or chemical bonds. 
They are able to entrap large volumes of water attracted by the high concentration of hydrophilic groups. 
When a dehydrated or deswollen hydrogel uptakes water, its polymer chains extend, creating a swelling pressure. 
For example, as reported in \cite{Wack_2009} weakly crosslinked poly(acrylic acid) (PAA) copolymers with polymer volume fractions between 0.03 and 0.30 exhibit a swelling pressure ranging from 0.20--4.23 MPa.
Polyelectrolyte hydrogels, which are carrying ionic groups on the comonomer units (like PAA), are able to reject salt ions from the solution, they absorb a solution of lower salinity than that they are equilibrated with.

@@

\paragraph{Physics behind the desalination process}
%The maximum entropy principle requires the density of charges to be more-less the same throughout the system medium. 
Since the polyelectrolyte gel has its own charges and its own neutralizing counterions, the density of mobile ions (which are able to freely enter and leave the gel) in gel appears to be smaller than their density outside the gel.
Therefore the internal solution of the gel has a lower density of mobile ions than that in the solution outside. 
In that sense, the gel acts as an osmotic membrane separating solutions \cite{Rud2020}. 
The driving force of the separation is the Donnan potential originated from the charges of the hydrogel network. 
The density of mobile ions difference between ions concentrations in the internal and external solutions
is defined by Donnan law \cite{Rud2018}
\begin{equation*}
\frac{\ccl\gel}{\cs}=\frac{\cs}{\cna\gel}=\sqrt{1+\left(\frac{\alpha}{2\cs\Vgel}\right)^{2}}\pm\frac{\alpha}{2\cs\Vgel}
\end{equation*}
where $\ccl\gel$ and $\cna\gel$ are the internal concentration of monovalent anions and cations, $\cs$ is the salinity of external solution,  $\Vgel$ is the gel molar volume (inverse density of gel segments) and $\alpha$ is ionization degree (in the case of our study $\alpha=1$). 
The $\pm$ sign in the formula accounts for the change between the cases of gel made of polyanion or polycation.

@@

The amount of the individual ion types in gel, $N_i$, is given by
\begin{equation}
N_i = N_i\gel + \cs\cdot(\Vbox - \Vgel)
\end{equation}
where $i \in \{\na, \cl\}$, $\Vgel$~--- volume of the gel, $\Vbox$~--- the volume in which the gel is compressed. 
Vbox$ is the gel volume in the free swelling equilibrium state.

\paragraph{Open system}
We simulate the open system using the grand reaction method \cite{Landsgesel2020a,Rud2020}.
This method is a hybrid of Molecular dynamics (MD) and Monte Carlo (MC). 
The whole simulation consists of MD and MC subsimulations, followed one by each other. 
The MD simulation is a standard Langevin dynamics \cite{Grest1986}, which models the mechanical motion of all particles in the system. 
MC simulates the thermodynamic equilibrium with reservoir and handles the ion exchange with the simulation box. 
The insertion (and deletion) of ion pairs is considered as a reaction of creation (or annihilation) of an ion pair 
\[
\varnothing\stackrel{K}{\leftrightarrows}\na+\cl
\]
with a reaction constant defined by the chemical potential of ions,
$K=\exp\left(\muna+\mucl\right).$

\paragraph{Closed system}
Analogical scheme simulates the gel in the closed system. 
The procedure represents a chain of subsequent MD and MC simulation, but now the molecular dynamics is running in two simulation boxes simultaneously.
The first box of the volume $\Vgel$ contains the gel with ions, and the second box, $\Vout$--- contains only ions. 
The total volume of both boxes is kept constant, $\Vbox=V_{\mathrm{gel}}+V_{\mathrm{out}}$, so the compression of the gel implies the decreasing $\Vgel$ and increasing $\Vout$.
These two boxes are represented in \reffig{fig:diamond}.

\paragraph{Algorithm}
The thermodynamic equilibrium between the two volumes implies the exchange of ion pairs between them, which is done by means of MC procedure in a way similar to described in \cite{Panagiotopoulos1988b}.
The simulation algorithm is following:
\begin{enumerate}
\item Simulate molecular dynamics in both boxes; collect the observables, for example, values of pressure $P$, to an array; 
\item Perform Monte-Carlo procedure simulating ion pair exchange; collect the arrays of observables samples: number of ions in both boxes, 
%$\nna\gel$,
$\Ncl\gel$ 
%$\nna\out$ 
and $\Ncl\out$ ; 
\item Repeat the procedure until the desired length of sample arrays is reached.
\end{enumerate}

@@

\paragraph{Compression in open system}
\begin{figure*}[t]
\subfloat[The gel partial pressure vs gel molar volume \label{fig: PV}]
{\includegraphics[width=0.48\textwidth]{figures/fig_PV_pub.pdf}}
\hspace{0.02\textwidth}
\subfloat[Supernate salinity vs gel molar volume\label{fig: CV}]
{\includegraphics[width=0.48\textwidth]{figures/fig_CV_pub.pdf}}
\caption{
The compression of the gel in open system (solid lines) and in closed system (dotted lines). 
Each solid curve corresponds to different salinity of the reservoir $\cs$ (see legend). Shadowed area limit the states with applied pressure below zero and above 5 bar.\label{fig: PV and CV}}
\end{figure*}
First, we run a set of simulations modeling the gel compression in \emph{open system}, \ie in equilibrium with a big bath of certain salinity, $\cs$. 
The simulations are run for a set of different simulation box volumes. 
Each simulation returns the averaged pressure, $P$, and the number of $\cl$ ions, which are present in the simulation box, $\ncl$.

@@

\paragraph{Compression in closed system}
%Each set of simulations in \emph{open system} defines the free swelling equilibrium states of the gel for each salinity, \ie the state where the gel partial pressure equals zero. 
The free swelling equilibrium states are characterized by the corresponding molar volume of the gel, $\Vgel^0$ and the amount of ions in gel \{$\Nna^0$, $\Ncl^0$\} 
(Index ``0'' stands for zero bar applied pressure).
We use these states as the starting ones for the gel compression in \emph{closed system}. 
Thus, we simulate the compression of the gel in a closed volume, $\Vbox$, which is equal to $\Vgel^0$, and contains $\Ncl^0$ of $\cl$ ions and the $\na$ ions in the amount neutralizing the system, $\Nna^0 = $$\Ncl^0 + \Ngel$.

@@

\paragraph{Desalination scheme}
As we have seen, the compression of the gel in a closed system affects the salinity, whereas the compression in an open system affects the amount of ions in a volume where the gel is compressed in. 
Here we show how to employ these phenomena for water desalination. 
The highlighted colored lines on plots of Figures \ref{fig: PV and CV} and \ref{fig: NV and CN} form a sequence of the following one by each other swellings and compressions of the hydrogel, correspondingly in \emph{open} and \emph{closed} systems. 
This sequence forms the water desalination process.
Starting from swelling the gel in the \emph{open} system at high salinity ($\cs=0.091$ mol/l, solid black line), the gel gets compressed in \emph{closed} system until the pressure reaches 5 bar (dashed black line). 
Then the same gel swells in a reservoir of a bit lower salinity (\ie $\cs=0.064$ mol/l, light blue line). 
After swelling, the gel is compressed again by the pressure of 5 bar in \emph{closed} system (dashed light blue line).  
Then the gel is put swelling in a reservoir of even smaller salinity ($\cs=0.045$ mol/l, solid yellow line). 
And so on. 
The chain of alternating swelling and compression steps ends up when salinity gets equal to $\cs=4\cdot10^{-3}$ mol/l after compression in closed system (dashed magenta line).
